Trends in Redwine by Neil Amin
========================================================

This dataset has 1599 records. It contains data on different red wines. The data
set has variables for fixed acidity, volatile acidity, citric acid, residual sug
ar, chlorides, free sulfur dioxide, total sulfur dioxide, density, ph, 
sulphates, and alcohol. The output is the quality of wine. My goal is to find 
trends between traits of wines that lead to a higher or lower quality wine.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code

library(ggplot2)
library(scales)
library(dplyr)
library(reshape2)
library(gridExtra)
library(lubridate)
library(swirl)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wineInfo <- read.csv("wineQualityReds.csv")
```

Here we can see basic information about the dataset. We get statistics about the
dataset and all 11 variables. X is not a variable but actually is the index of 
each record in the data. Quality is the output that we will be trying to find 
trends about.

# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}
str(wineInfo)
names(wineInfo)
summary(wineInfo)
```

```{r echo=FALSE}
hist1 <- qplot(x = quality, data = wineInfo, bins=16, xlab = "Quality", 
               ylab = "Count", main = "Histogram of Quality")

hist2 <- qplot(x = fixed.acidity, data = wineInfo, bins=16, 
               xlab = "Fixed Acidity", ylab = "Count",
               main = "Histogram of Fixed Acidity")

hist3 <- qplot(x = volatile.acidity, data = wineInfo, bins=12, 
               xlab = "Volatile Acidity", ylab = "Count", 
               main = "Histogram of Volatile Acidity")

hist4 <- qplot(x = citric.acid, data = wineInfo, bins=20, xlab = "Citric Acid", 
               ylab = "Count", main = "Histogram of Citric Acid")

hist5 <- qplot(x = residual.sugar, data = wineInfo, bins=20, 
               xlab = "Residual Sugar", ylab = "Count", 
               main = "Histogram of Residual Sugar")

hist6 <- qplot(x = chlorides, data = wineInfo, bins=20, xlab = "Chlorides", 
               ylab = "Count", main = "Histogram of Chlorides")

hist7 <- qplot(x = free.sulfur.dioxide, data = wineInfo, bins=15,
               xlab = "Free Sulfur Dioxide", ylab = "Count", 
               main = "Histogram of Free Sulfur Dioxide")

hist8 <- qplot(x = total.sulfur.dioxide, data = wineInfo, bins=40, 
               xlab = "Total Sulfur Dioxide", ylab = "Count", 
               main = "Histogram of Total Sulfur Dioxide")

hist9 <- qplot(x = density, data = wineInfo, bins=30, xlab = "Density", 
               ylab = "Count", main = "Histogram of Density")

hist10 <- qplot(x = pH, data = wineInfo, bins=30, xlab = "pH", 
                ylab = "Count", main = "Histogram of pH")

hist11<- qplot(x = sulphates, data = wineInfo, bins=30, xlab = "Sulfates", 
               ylab = "Count", main = "Histogram of Sulphates")

hist12<- qplot(x = alcohol, data = wineInfo, bins=30, xlab = "Alcohol", 
               ylab = "Count", main = "Histogram of Alcohol")

grid.arrange(hist1,hist2,hist3,hist4, ncol = 2)
grid.arrange(hist5,hist6,hist7,hist8, ncol = 2)
grid.arrange(hist9,hist10,hist11,hist12, ncol = 2)

```

These are historgrams for all 11 variables and the output (quality). All of the 
graphs appear to have a somewhat normal distribution except for residual sugar, 
chlorides, free sulfur dioxide, and total sulfur dioxide. If we limit the 
x values then we won't have this problem.

```{r, echo=FALSE}
hist5 <- qplot(x = residual.sugar, data = wineInfo, bins=20, 
               xlab = "Residual Sugar", ylab = "Count", 
               main = "Limited Histogram of Residual Sugar") + xlim(0, 5)

hist6 <- qplot(x = chlorides, data = wineInfo, bins=20, xlab = "Chlorides", 
               ylab = "Count", main = "Limited Histogram of Chlorides") + 
               xlim(0, .2)

hist7 <- qplot(x = free.sulfur.dioxide, data = wineInfo, bins=15, 
               xlab = "Free Sulfur Dioxide", ylab = "Count", 
               main = "Limited Free Sulfur Dioxide") + xlim(0, 30)

hist8 <- qplot(x = total.sulfur.dioxide, data = wineInfo, bins=10, 
               xlab = "Total Sulfur Dioxide", ylab = "Count", 
               main = "Limited Total Sulfur Dioxide") + xlim(0, 70)

grid.arrange(hist5,hist6,hist7,hist8, ncol = 2)
```

If we alter the bin size and the x values that are included we can see more 
normalized graph. This gets rid of the smaller, outlier values so that we can 
see the main bulk of records.

```{r, echo=FALSE}
g <- ggplot(wineInfo, aes(quality))
g + geom_bar()
```

With this graph, we can see the majority of the wine falls between quality 5 
and 6. This lets gives us the indicator that anything above a 6 in quality is 
better than average wine. We want to look for trends in what kinds of wine have 
this better than average rating.

# Univariate Analysis

Since we are investigating what characteristics of the wine make a high quality 
wine. we are going to focus more on the subset of the data with a higher 
quality. We will also use the subset of low quality wine which is below a 
quality rating of 3 in order to see if it follows opposite trends which would 
indicate some correlations. Most of our data was normally distributes from the 
beginning but it is important that we find the normal distribution of the 
others by removing the outliers.

# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

box1 <- ggplot(wineInfo) + 
        geom_boxplot(aes(x=factor(quality), y=fixed.acidity)) + 
        labs(title = "Fixed Acidity Box Plot") + xlab("Quality") + ylab("Fixed Acidity")

box2 <- ggplot(wineInfo) + 
        geom_boxplot(aes(x=factor(quality), y=volatile.acidity)) + 
        labs(title = "Volatile Acidity Box Plot") + xlab("Quality") + ylab("Volatile Acidity")

box3 <- ggplot(wineInfo) + 
        geom_boxplot(aes(x=factor(quality), y=citric.acid)) + 
        labs(title = "Citric Acid Box Plot") + xlab("Quality") + ylab("Citric Acid")

box4 <- ggplot(wineInfo) + 
        geom_boxplot(aes(x=factor(quality), y=residual.sugar)) + 
        labs(title = "Residual Sugar Box Plot") + xlab("Quality") + ylab("Residual Sugar")

box5 <- ggplot(wineInfo)+ 
        geom_boxplot(aes(x=factor(quality), y=chlorides)) + 
        labs(title = "Chlorides Box Plot") + xlab("Quality") + ylab("Chlorides")

box6 <- ggplot(wineInfo) + 
        geom_boxplot(aes(x=factor(quality), y=free.sulfur.dioxide)) + 
        labs(title = "Sulfur Dioxide Box Plot") + xlab("Quality") + ylab("Sulfur Dioxide")

box7 <- ggplot(wineInfo) + 
        geom_boxplot(aes(x=factor(quality), y=total.sulfur.dioxide)) + 
        labs(title = "Total Sulfur Dioxide Box Plot") + xlab("Quality") + ylab("Total Sulfur Dioxide")

box8 <- ggplot(wineInfo) + 
        geom_boxplot(aes(x=factor(quality), y=density)) + 
        labs(title = "Density Box Plot") + xlab("Quality") + ylab("Density")

box9 <- ggplot(wineInfo) + 
        geom_boxplot(aes(x=factor(quality), y=pH)) + 
        labs(title = "pH Box Plot") + xlab("Quality") + ylab("pH")

box10 <- ggplot(wineInfo) + 
         geom_boxplot(aes(x=factor(quality), y=sulphates)) + 
         labs(title = "Sulphates Box Plot") + xlab("Quality") + ylab("Sulphates")

box11 <- ggplot(wineInfo) + 
         geom_boxplot(aes(x=factor(quality), y=alcohol))  + 
         labs(title = "Alcohol Box Plot") + xlab("Quality") + ylab("Alcohol(%)")

grid.arrange(box1,box2,box3,box4+ coord_trans(y = "log10"),ncol=2)
cor.test(wineInfo$quality, wineInfo$fixed.acidity)
cor.test(wineInfo$quality, wineInfo$volatile.acidity)
cor.test(wineInfo$quality, wineInfo$citric.acid)
cor.test(wineInfo$quality, wineInfo$residual.sugar)
```

The first box plot compares Fixed Acidity vs Quality. I wanted to see if there 
was a trend between levels of fixed acidity and quality. From looking at these 
plots it doesn't seem that there is much of a trend.

The second box plot compares Volatile Acidity vs Quality. There appears to be a 
trend that less volatile acidity tends to have better quality.

The third box plot compares Citric Acid vs Quality. This plot trends the 
opposite way of volatile acidity in that more citric acid tended to get better 
ratings.

The fourth plot compares residual sugar with quality. It appears that there is 
not much trend between residual sugar and quality.

```{r, echo=FALSE}
grid.arrange(box5 + coord_trans(y = "log10"),box6,box7,box8,ncol=2)
cor.test(wineInfo$quality, wineInfo$chlorides)
cor.test(wineInfo$quality, wineInfo$free.sulfur.dioxide)
cor.test(wineInfo$quality, wineInfo$total.sulfur.dioxide)
cor.test(wineInfo$quality, wineInfo$density)
```

The first box plot here compares chlorides to quality. Here it seems that less 
chlorides might translate to better quality. This is not 100% clear so more 
investigation will be needed.

The second box compares sulfur dioxide to quality. There doesn't appear to much 
correlation here.

The third box plot is total sulfur dioxide vs quality. There also doesn't
appear to be much correlation here.

The fourth box plot is density vs quality. It seems there might be a 
correlation to less dense wine being higher quality but like chlorides there 
needs to be more investigation.

```{r, echo=FALSE}
grid.arrange(box9,box10,box11, ncol=2)
cor.test(wineInfo$quality, wineInfo$sulphates)
cor.test(wineInfo$quality, wineInfo$pH)
cor.test(wineInfo$quality, wineInfo$alcohol)
```

This first plot is pH vs quality. There doesn't appear to be much correlation.

The second plot is sulphates vs quality. It appears that higher sulphates 
translate to higher quality wine.

The last plot is alcohol vs quality. There is a strong trend between higher 
alcohol and higher quality.

After conducting Pearson's correlation on all of the data it is clear that 
alcohol, fixed acidity, citric acid, and sulphates all have positive 
correlations with quality. This is in contrast to density, total sulfur dioxide, 
free sulfur, chlorides, total sulfur,and volatile acidity which were negative.

```{r, echo=FALSE}
ggplot(aes(x = fixed.acidity, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = volatile.acidity, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = citric.acid, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = residual.sugar, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = chlorides, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = free.sulfur.dioxide, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = total.sulfur.dioxide, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = density, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = pH, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = sulphates, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')

ggplot(aes(x = alcohol, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')
```

These graphs show the linear regression line when the characteristics are 
compared to quality. The lines that had the largest slopes were the ones that 
also happened to have the largest positve correlations. 

# Bivariate Analysis

My first goal was to use the box plots to figure out what variables had a 
strong amount of high quality wines. It seemed that a certain few 
haracteristics (alcohol, fixed acidity, citric acid, and sulphates). I decided 
to futhur investigate by checking the correlations. It seems the same 
characteristics had high positive correlations. Next I plotted all of the 
variables against quality and the ones with the highest slopping regression 
lines were the same as previously. These few charactertisitcs are going to
require more investigation by me.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
#bins for alcohol consentration 
wineInfo$alc.bucket <- cut(wineInfo$alcohol, c(8,11,14))



ggplot(aes(x=volatile.acidity, y= quality),
       data = subset(wineInfo, !is.na(alc.bucket))) +
  geom_point(aes(color = alc.bucket),
            stat = 'summary', fun.y = mean) +
  geom_smooth(method = 'lm')

ggplot(aes(x=citric.acid, y= quality),
       data = subset(wineInfo, !is.na(alc.bucket))) +
  geom_point(aes(color = alc.bucket),
            stat = 'summary', fun.y = mean) +
  geom_smooth(method = 'lm')

ggplot(aes(x=sulphates, y= quality),
       data = subset(wineInfo, !is.na(alc.bucket))) +
  geom_point(aes(color = alc.bucket),
            stat = 'summary', fun.y = mean) +
  geom_smooth(method = 'lm') 

```

Because alcohol had the strongest correlation, I took my earlier graphs and 
compared them again with alcohol as a 2nd variable. The dots are shaded 
different based on the different levels of alcohol in each. It seems that this 
graph definitely indicates that higher alcohol is definitely correlated with 
higher quality.

# Multivariate Analysis

It is clear from the multivariate analysis that alcohol was the most important 
factor in determining quality. The other three characteristics all seemed to 
have a smaller porportional effect on the quality as well.

# Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One}
g2 <- ggplot(wineInfo, aes(quality)) + 
      labs(x="Quality", y="Count", title="Quality Count")
g2 + geom_bar()
```

### Description One

I used this graph because it gave me a good initial indication as to what
would be considered good wine vs average wine. I noticed that most of the 
wine feel into a rating of 5,or 6 which meant that anything about 6 could 
be considered high quality. This bench mark was used for much of my analysis.


### Plot Two

```{r echo=FALSE, Plot_Two}
box11
```

### Description Two

I chose to include the box plot of alcohol and quality because I felt that it 
was the first piece of information that caught my eye when trying to find 
characteristics of high quality wine. The box plots for the higher quality 
wines tended to fall higher in alcohol content.

### Plot Three

```{r echo=FALSE, Plot_Three}

ggplot(aes(x = sulphates, y = quality), data = wineInfo) + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')+
  labs(x="Sulphates(mg/l)", y="Quality", title="Quality vs Sulphates")
```

### Description Three

I chose to add this graph because sulfates had the second highest positive 
correlation with quality. This gave a look at where high alcohol and high 
sulfate containing wine ranked in the quality spectrum.


# Reflection

I decided to investigate what characteristics of the wine make a high quality 
wine. After figuring out what a strong rating was using a histogram for 
quality, my first goal was to use the box plots to figure out what variables 
had a strong amount of high quality wines. A certain few characteristics 
(alcohol, fixed acidity, citric acid, and sulphates) stuck out from others 
based on their plots. I then futhur investigated by checking the correlations 
of the 11 variables vs quality. I was able to conclude that same 
characteristics had high positive correlations. 

Finally I plotted all of the variables against quality to confirm my findings 
and once again the variables with the highest slopping regression lines were 
the same as previously. While looking at these, alcohol in particular also had 
the strongest trend so I focused in on that by adding alcohol level to all of 
the points in the other 3 graphs. It became clear from the multivariate 
analysis that alcohol was the most important factor in determining quality. 
The other three characteristics all seemed to have a smaller porportional 
effect on the quality as well.

My biggest obstacle was removing outliers from the data so that I could 
normalize it. Residual sugar, chlorides, free sulfur dioxide, and total 
sulfur dioxide were the characteristics that needed to have outliers removed. 
Another problem I encountered was being able to create box plots with an 
x value that was continous. I was able to fix this using the group function.

Going forward I think there needs to be an investigation to whether alcohol 
is causing a random correlation or an actual causation to an increased quality 
level. I think BAC of the person creating the quality rating should be 
accounted for if they are making a judgement based on drinking the wine.